---
title: "Plot geom_sf() On OpenStreetMap Tiles"
author: Hiroaki Yutani
date: "2018-06-09"
categories: ["ggplot2", "GIS"]
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(dev = "ragg_png")
```

[mapview](https://r-spatial.github.io/mapview/) is a very nice package to explore an `sf` object. It can overlay `sf` object on the map images:

```{r mapview}
nc <- sf::read_sf(system.file("shape/nc.shp", package="sf"))

# mapview::mapview(nc)
```

But, how can I do this with ggplot2? (My friend told me `mapview::mapshot()` can generate a PNG, but I want to do this with ggplot2!)

## RTFM

Before anything, I need to read [Tile Usage Policy](https://operations.osmfoundation.org/policies/tiles/) to use the OpenStreetMap tiles. For "Requirements" section, this is important:

> Clearly display license attribution.

According to [Copyright and License](https://www.openstreetmap.org/copyright), it's so simple as just adding this caption to my plots:

```{r caption, eval=FALSE}
labs(caption = "\U00a9 OpenStreetMap contributors")
```

For "Technical Usage Requirements" section, I have to read this more carefully. Let's look at the requirements one by one.

> Valid HTTP User-Agent identifying application. Faking another app's User-Agent WILL get you blocked.

Oh, it seems I need to add `User-Agent` header. OK, let's invent some nice name... If I use `httr::GET()`, the code will probably like this:

```{r user-agent, eval=FALSE}
GET(
  "https://...",
  add_headers(`User-Agent` = "Yutani's blog post")
)
```

> If known, a valid HTTP Referer.

I don't have `Referer`s, so I skip this.

> DO NOT send no-cache headers. ("Cache-Control: no-cache", "Pragma: no-cache" etc.)

I do nothing other than swearing I'll never use this header.

> Cache Tile downloads locally according to HTTP Expiry Header, alternatively a minimum of 7 days.

Ah, this is important. Let's implement later.

> Maximum of 2 download threads. (Unmodified web browsers' download thread limits are acceptable.)

Fortunately, I'm not good at parallelism, so this is fine.

## Get Tile URLs

According to [Slippy map tilenames](https://wiki.openstreetmap.org/wiki/Slippy_map_tilenames#Tile_servers), the URL of a tile follows this format:

    https://[abc].tile.openstreetmap.org/zoom/x/y.png 

I have to fill these four parts:

-   `[abc]`
-   `zoom`
-   `x`
-   `y`

Let's look at these one by one.

### `[abc]`

`[abc]` means there are three domains; `a.tile.openstreetmap.org`, `b.tile.openstreetmap.org`, `c.tile.openstreetmap.org`. But, why? It says:

> Browser-based applications can thus request multiple tiles from multiple subdomains faster than from one subdomain.

So, as I'm not browser-based, I can choose arbitrary one.

### `zoom`

Zoom parameter is an integer number from 0 to 19. If `zoom` is 0, there's only one tile. Likewise 2 x 2 tiles for zoom `1`, 4 x 4 tiles for zoom `2`, and so on. Then, which one should I choose? This can be roughly determined based on the size of the bbox (boundary box) of the `sf` object.

```{r bbox}
# get the bbox
b <- sf::st_bbox(nc)
b

# calculate the lengths of x and y of the bbox
x_len <- b["xmax"] - b["xmin"]
y_len <- b["ymax"] - b["ymin"]

# calculate the minimum zoom level that is smaller than the lengths
x_zoom <- sum(x_len < 360 / 2^(0:19)) - 1
y_zoom <- sum(y_len < 170.1022 / 2^(0:19)) - 1

zoom <- min(x_zoom, y_zoom)
zoom
```

But, since the tile is so small as 256 × 256 pixel, it's often better to zoom more.

```{r bbox-zoomin}
zoom <- zoom + 2
```

(I'm not sure how to do this correctly, I guess the zoom level should be determined by the size of the plot canvas, not the bbox.)

### `x` and `y`

A pair of `x` and `y` represents the location of a tile. `x` corresponds to longitude, `y` to latitude. If the `zoom` is given, I can convert longitudes and latitudes to `x` and `y` according to the [pseudo code on OpenStreetMap's wiki](https://wiki.openstreetmap.org/wiki/Slippy_map_tilenames#Pseudo-code):

```{r calculate_x_y}
sec <- function(x) {
  1 / cos(x)
}

lonlat2xy <- function(lat_deg, lon_deg, zoom) {
  n <- 2^zoom

  x <- (n * (lat_deg + 180)) %/% 360
  lon_rad <- lon_deg * pi / 180
  y <- (n * (1 - log(tan(lon_rad) + sec(lon_rad)) / pi)) %/% 2

  list(x = x, y = y)
}
```

But, how can I find the set of tiles which covers the whole bbox?

```{r bbox-plot}
library(ggplot2)

p <- ggplot(nc) +
  geom_sf() +
  annotate("rect", xmin = b["xmin"], xmax = b["xmax"], ymin = b["ymin"], ymax = b["ymax"],
           colour = alpha("red", 0.4), fill = "transparent", linetype = "dashed", size = 1.2)
p
```

This problem can be simplified; I can focus only two corners, the north-west and the south-east. If I calculate which tiles of `x` and `y` those two points fall in, I can find the rest of the tiles by filling the sequences of `x` and `y` between these two tiles.

```{r plot-corner}
corners <- expand.grid(x = b[c(1, 3)], y = b[c(2, 4)])

p +
  geom_point(aes(x, y), corners[2:3,], colour = "red", size = 5)
```

Here's the tiles:

```{r calculate-tiles}
xy <- lonlat2xy(b[c("xmin", "xmax")], b[c("ymin", "ymax")], zoom)

tiles <- expand.grid(x = seq(xy$x["xmin"], xy$x["xmax"]),
                     y = seq(xy$y["ymin"], xy$y["ymax"]))

tiles
```

### Tile URLs

From the results above, I can yield the URLs of the tiles:

```{r urls}
urls <- sprintf("https://a.tile.openstreetmap.org/%d/%d/%d.png", zoom, tiles$x, tiles$y)
urls
```

OK, now I can download them. But, before that, let's confirm that the tiles really cover the expected area.

### Tile positions

To plot these tiles, I calculate the north-west corner of a tile by the following code (the pseudo code for this is also found on the wiki):

```{r xy2lonlat}
xy2lonlat <- function(x, y, zoom) {
  n <- 2^zoom

  lon_deg <- x / n * 360.0 - 180.0
  lat_rad <- atan(sinh(pi * (1 - 2 * y / n)))
  lat_deg <- lat_rad * 180.0 / pi

  list(lon_deg = lon_deg, lat_deg = lat_deg)
}
```

Then, south-east corners can be also calculated easily. Let's calculate the both corners and bind them to a data.frame.

```{r tile-positions}
library(purrr)
library(dplyr, warn.conflicts = FALSE)

nw_corners <- pmap_dfr(tiles, xy2lonlat, zoom = zoom)
# add 1 to x and y to get the south-east corners
se_corners <- pmap_dfr(mutate_all(tiles, `+`, 1), xy2lonlat, zoom = zoom)

names(nw_corners) <- c("xmin", "ymax")
names(se_corners) <- c("xmax", "ymin")

tile_positions <- bind_cols(nw_corners, se_corners)
tile_positions
```

Now I can plot the empty tiles as below:

```{r plot-empty-tiles}
p +
  geom_point(aes(x, y), corners[2:3,], colour = "red", size = 5) +
  geom_rect(data = tile_positions,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            colour = "blue", fill = "transparent")
```

Yay, confirmed! Let's proceed to the next step.

## Get tile data

If I just get the response from a URL, `httr::GET()` is handy. But, this time, for the requirement of caching, I have to save the responses to disk first. So, I use `curl::curl_download()` here.

Note that PNG data can be read into R session by `png::readPNG()`.

```{r curl-get-data}
get_tile <- function(url) {
  # build a local path
  path <- stringr::str_extract(url, "/\\d+/\\d+/\\d+.png")
  local_png <- here::here(file.path("data", "osm-tiles", path))

  if (!file.exists(local_png)) {
    dir.create(dirname(local_png), showWarnings = FALSE, recursive = TRUE)
    
    # add header
    h <- curl::new_handle()
    curl::handle_setheaders(h, `User-Agent` = "Yutani's blog post")
    
    curl::curl_download(url, destfile = local_png)
  }

  png::readPNG(local_png)
}
```

Then, let's get all tiles.

```{r curl-do-download}
pngs <- map(urls, get_tile)
```

## Plot tiles

To plot tiles, I use `annotation_raster()`, whose necessary arguments are:

-   `raster`
-   `xmin`
-   `xmax`
-   `ymin`
-   `ymax`

The first one is `pngs` and the others are contained in `tile_positions`. So, let's combine them so that I can use `pmap()`.

```{r annotation-args}
args <- tile_positions %>%
  mutate(raster = pngs)

args
```

Now I can plot tiles at last.

```{r plot, preview=TRUE}
ggplot(nc) +
  pmap(args, annotation_raster, interpolate = TRUE) +
  geom_sf(fill = alpha("red", 0.3)) +
  # don't forget the license notice!
  labs(caption = "\U00a9 OpenStreetMap contributors") +
  theme_minimal()
```

Done!

But, I didn't expect the code would be this long... Maybe I need to create a package for this.

## Caveats

<strike>Note that, I didn't care about the CRS because `nc`'s CRS is fortunately EPSG 3857, which OpenStreetMap uses. If the `sf` object I want to plot has the different CRS, there may be a bit more to consider (and I don't understand the CRS well...).</strike>

**Update:**

Sorry, I was wrong... `nc`'s CRS is EPSG 4267 and OpenStreetMap tiles use EPSG 4326. Thanks Edzer for [pointing this out](https://twitter.com/edzerpebesma/status/1005494350585978893)!

```{r nc_crs}
sf::st_crs(nc)
```

So, I should have converted `nc` to the CRS first.

```{r nc_transform}
nc_4326 <- sf::st_transform(nc, 4326)
```

Fortunately, the difference between EPSG 4267 and EPSG 4326 is rather negligible for this scale, so the result map should look almost same if I used `nc_4326` instead of `nc`. Here's the difference (can you see there's a little red colors?):

```{r difference}
nc_4326_not_transformed <- sf::`st_crs<-`(nc, 4326)

ggplot() +
  geom_sf(data = nc_4326_not_transformed, fill = "transparent", colour = "red") +
  geom_sf(data = nc_4326, fill = "transparent", colour = "blue") +
  theme_minimal()
```
