---
title: gghighlight 0.2.0
description: |
  gghighlight 0.2.0 is released!
author: 'Hiroaki Yutani'
date: '2020-02-17'
categories:
  - gghighlight
  - ggplot2
  - R package
---

[gghighlight 0.2.0](https://yutannihilation.github.io/gghighlight/news/index.html#gghighlight-020) is on CRAN a while ago. This post briefly introduces the three new features. For basic usages, please refer to "[Introduction to gghighlight](https://yutannihilation.github.io/gghighlight/articles/gghighlight.html)".

## `keep_scales`

To put it simply, gghighlight doesn't drop any data points but drops their colours. This means, while non-colour scales (e.g. `x`, `y` and `size`) are kept as they are, colour scales get shrinked. This might be inconvenient when we want to compare the original version and the highlighted version, or the multiple highlighted versions.

```{r keep_scale}
library(gghighlight)
library(patchwork)

set.seed(3)

d <- data.frame(
  value = 1:9,
  category = rep(c("a","b","c"), 3),
  cont_var = runif(9),
  stringsAsFactors = FALSE
)

p <- ggplot(d, aes(x = category, y = value, color = cont_var)) +
  geom_point(size = 10) +
  scale_colour_viridis_c()

p1 <- p + ggtitle("original")
p2 <- p + 
  gghighlight(dplyr::between(cont_var, 0.3, 0.7),
              use_direct_label = FALSE) +
  ggtitle("highlighted")

p1 * p2
```

You can see the colour of the points are different between the left plot and the right plot because the scale of the colours are different. In such a case, you can specify `keep_scale = TRUE` to keep the original scale (under the hood, gghighlight simply copies the original data to `geom_blank()`).

```{r keep_scale2, preview=TRUE}
p3 <- p +
  gghighlight(dplyr::between(cont_var, 0.3, 0.7),
              keep_scales = TRUE,
              use_direct_label = FALSE) +
  ggtitle("highlighted (keep_scale = TRUE)")

p1 * p3
```

## `calculate_per_facet`

When used with `facet_*()`, `gghighlight()` puts unhighlighted data on all facets and calculate the predicates on the whole data.

```{r calculate_per_facet1}
Sys.setlocale(locale = "C")
set.seed(16)

d <- tibble::tibble(
  day = rep(as.Date("2020-01-01") + 0:89, times = 4),
  month = lubridate::ceiling_date(day, "month"),
  value = c(
    cumsum(runif(90, -1.0, 1.0)),
    cumsum(runif(90, -1.1, 1.1)),
    cumsum(runif(90, -1.1, 1.0)),
    cumsum(runif(90, -1.0, 1.1))
  ),
  id = rep(c("a", "b", "c", "d"), each = 90)
)

p <- ggplot(d) +
  geom_line(aes(day, value, colour = id)) +
  facet_wrap(~ month, scales = "free_x")

p + 
  gghighlight(mean(value) > 0, keep_scales = TRUE)
```

But, it sometimes feels better to highlight facet by facet. For such a need, `gghighlight()` now has a new argument `calculate_per_facet`.

```{r calculate_per_facet2}
p + 
  gghighlight(mean(value) > 0,
              calculate_per_facet = TRUE,
              keep_scales = TRUE)
```

Note that, as a general rule, only the layers **before** adding `gghighlight()` are modified. So, if you add facet\_\*() after adding gghighlight(), this option doesn't work (though this behaviour might also be useful in some cases).

```{r calculate_per_facet3}
ggplot(d) +
  geom_line(aes(day, value, colour = id)) +
  gghighlight(mean(value) > 0,
              calculate_per_facet = TRUE,
              keep_scales = TRUE) +
  facet_wrap(~ month, scales = "free_x")
```

## `unhighlighted_params`

`gghighlight()` now allows users to override the parameters of unhighlighted data via `unhighlighted_params`. This idea was suggested by [\@ClausWilke](https://twitter.com/ClausWilke/status/1014529225402003456).

<blockquote class="twitter-tweet">

<p lang="en" dir="ltr">

I *think* you could support a broader set of use cases if you allowed a list of aesthetics default values, like `bleach_aes = list(colour = &quot;grey40&quot;, fill =&quot;grey80&quot;, size = 0.2)`.

</p>

--- Claus Wilke (@ClausWilke) <a href="https://twitter.com/ClausWilke/status/1014529225402003456?ref_src=twsrc%5Etfw">July 4, 2018</a>

</blockquote>

```{=html}
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
```
To illustrate the original motivation, let's use an example on [the ggridges' vignette](https://cran.r-project.org/web/packages/ggridges/vignettes/gallery.html). gghighlight can highlight almost any Geoms, but it doesn't mean it can "unhighlight" arbitrary colour aesthetics automatically. In some cases, you need to unhighlight them manually. For example, `geom_density_ridges()` has `point_colour`.

```{r unhighlighted_params1}
library(ggplot2)
library(gghighlight)
library(ggridges)

p <- ggplot(Aus_athletes, aes(x = height, y = sport, color = sex, point_color = sex, fill = sex)) +
  geom_density_ridges(
    jittered_points = TRUE, scale = .95, rel_min_height = .01,
    point_shape = "|", point_size = 3, size = 0.25,
    position = position_points_jitter(height = 0)
  ) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0), name = "height [cm]") +
  scale_fill_manual(values = c("#D55E0050", "#0072B250"), labels = c("female", "male")) +
  scale_color_manual(values = c("#D55E00", "#0072B2"), guide = "none") +
  scale_discrete_manual("point_color", values = c("#D55E00", "#0072B2"), guide = "none") +
  coord_cartesian(clip = "off") +
  guides(fill = guide_legend(
    override.aes = list(
      fill = c("#D55E00A0", "#0072B2A0"),
      color = NA, point_color = NA)
    )
  ) +
  ggtitle("Height in Australian athletes") +
  theme_ridges(center = TRUE)

p + 
  gghighlight(sd(height) < 5.5)
```

You should notice that these vertical lines still have their colours. To grey them out, we can specify `point_colour = "grey80"` on `unhighlighted_params` (Be careful, `point_color` doesn't work...).

```{r unhighlighted_params2}
p + 
  gghighlight(sd(height) < 5.5, 
              unhighlighted_params = list(point_colour = "grey80"))
```

`unhighlighted_params` is also useful when you want more significant difference between the highlighted data and unhighligted ones. In the following example, `size` and `colour` are set differently.

```{r unhighlighted_params3}
set.seed(2)
d <- purrr::map_dfr(
  letters,
  ~ data.frame(
      idx = 1:400,
      value = cumsum(runif(400, -1, 1)),
      type = .,
      flag = sample(c(TRUE, FALSE), size = 400, replace = TRUE),
      stringsAsFactors = FALSE
    )
)

ggplot(d) +
  geom_line(aes(idx, value, colour = type), size = 5) +
  gghighlight(max(value) > 19,
              unhighlighted_params = list(size = 1, colour = alpha("pink", 0.4)))
```
