---
title: "dplyr::select() Accepts Characters Since Version 0.7.0"
date: "2017-10-25"
categories: ["R"]
tags: ["dplyr", "tidyeval"]
---

Some of my friends didn't aware that dplyr now accepts characters. Did you?

For example, this expression

```{r select-nse, eval=FALSE}
select(iris, Sepal.Length, Petal.Length)
```

can be also written in this way:

```{r select-character, eval=FALSE}
select(iris, "Sepal.Length", "Petal.Length")
```

or in this way:

```{r select-character2, eval=FALSE}
select(iris, c("Sepal.Length", "Petal.Length"))
```

For the semantics of `select()`, you can find a good explanation in [the vignette](http://dplyr.tidyverse.org/articles/dplyr.html#selecting-operations).

## Tidyeval?

You may want to write some code that selects columns programmatically using variables.

```{r select-danager, eval=FALSE}
x <- c("col1", "col2")
select(some_data, x)
```

But, be fareful, `x` can be ambiguous.
If `some_data` has a column named `x`, it selects the column `x` instead of `col1` and `col2`, the content of the variable `x`.

To avoid these troubles, you should explicitly unquote `x` by `!!` or `!!!` (both are ok in this case).

```{r select-tidyeval, eval=FALSE}
x <- c("col1", "col2")
select(some_data, !! x)
```

`rlang::expr()` can show how the expression is unquoted.

```{r select-tidyeval2}
x <- c("col1", "col2")

rlang::expr(select(some_data, !! x))

rlang::expr(select(some_data, !!! x))
```
