---
title: "Introduction to gghighlight: Highlight ggplot's Lines and Points with Predicates"
date: "2017-10-06"
categories: ["R"]
tags: ["gghighlight", "ggplot", "package"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

Suppose we have a data that has too many series like this:

```{r set-data}
set.seed(2)
d <- purrr::map_dfr(
  letters,
  ~ data.frame(idx = 1:400,
               value = cumsum(runif(400, -1, 1)),
               type = .,
               stringsAsFactors = FALSE))
```

For such data, it is almost impossible to identify a series by its colour as their differences are so subtle.

```{r plot}
library(ggplot2)

ggplot(d) +
  geom_line(aes(idx, value, colour = type))
```

Highlight lines with ggplot2 + dplyr
------------------------------------

So, I am motivated to filter data and map colour only on that, using dplyr:

```{r dplyr}
library(dplyr, warn.conflicts = FALSE)

d_filtered <- d %>%
  group_by(type) %>% 
  filter(max(value) > 20) %>%
  ungroup()

ggplot() +
  # draw the original data series with grey
  geom_line(aes(idx, value, group = type), data = d, colour = alpha("grey", 0.7)) +
  # colourise only the filtered data
  geom_line(aes(idx, value, colour = type), data = d_filtered)
```

But, what if I want to change the threshold in predicate (`max(.data$value) > 20`) and highlight other series as well? It's a bit tiresome to type all the code above again every time I replace `20` with some other value.

Highlight lines with gghighlight
-------------------------------

**gghighlight** package provides two functions to do this job. You can install this via [CRAN](https://cran.r-project.org/package=gghighlight) (or [GitHub](https://github.com/yutannihilation/gghighlight/))

```{r gh-installation, eval = FALSE}
install.packages("gghighlight")
```

`gghighlight_line()` is the one for lines. The code equivalent to above (and more) can be this few lines:

```{r gghighlight-line-basic}
library(gghighlight)

gghighlight_line(d, aes(idx, value, colour = type), predicate = max(value) > 20)
```

As `gghighlight_*()` returns a ggplot object, it is fully customizable just as we usually do with ggplot2 like custom themes and facetting.

```{r gghighlight-theme}
library(ggplot2)

gghighlight_line(d, aes(idx, value, colour = type), max(value) > 20) +
  theme_minimal()
```

```{r gghighlight-facet}
gghighlight_line(d, aes(idx, value, colour = type), max(value) > 20) +
  facet_wrap(~ type)
```

By default, `gghighlight_line()` calculates `predicate` per group, more precisely, `dplyr::group_by()` + `dplyr::summarise()`. So if the predicate expression returns multiple values per group, it ends up with an error like this:

```{r grouped-error, error=TRUE}
gghighlight_line(d, aes(idx, value, colour = type), value > 20)
```


Highlight points with gghighlight
---------------------------------

`gghighlight_point()` highlight points. While `gghighlight_line()` evaluates `predicate` by grouped calculation (`dplyr::group_by()`), by default, this function evaluates it by ungrouped calculation.

```{r gghighlight-point}
set.seed(19)
d2 <- sample_n(d, 100L)

gghighlight_point(d2, aes(idx, value), value > 10)
```

As the job is done without grouping, it's better to provide `gghighlight_point()` a proper key for label, though it tries to choose proper one automatically. Specifying `label_key = type` will stop the warning above:

```{r gghighlight-point-nowarn, eval=FALSE}
gghighlight_point(d2, aes(idx, value), value > 10, label_key = type)
````

You can control whether to do things with grouping by `use_group_by` argument. If this set to `TRUE`, `gghighlight_point()` evaluate `predicate` by grouped calculation.

```{r gghighlight-point-grouped}
gghighlight_point(d2, aes(idx, value, colour = type), max(value) > 15, label_key = type,
                  use_group_by = TRUE)
```

Non-logical predicate
---------------------

(Does "non-logical predicate" make sense...? Due to my poor English skill, I couldn't come up with a good term other than this. Any suggestions are wellcome.)

By the way, to construct a predicate expression like bellow, we need to determine a threshold (in this example, `20`). But it is difficult to choose a nice one before we draw plots. This is a chicken or the egg situation.

```{r predicate-example, eval=FALSE}
max(value) > 20
```

So, `gghiglight_*()` allows predicates that will be evaluated into non-logical values. The result value will be used to sort data, and the top `max_highlight` data points/series will be highlighted. For example:

```{r non-logical-predicate}
gghighlight_line(d, aes(idx, value, colour = type), predicate = max(value),
                 max_highlight = 6)
```

Caveats
-------

Seems cool? gghighlight is good to explore data by changing a threshlold little by little. But, the internals are not so efficient, as it does almost the same calculation everytime you execute `gghighlight_*()`, which may get slower when it works with larger data. Consider doing this by using vanilla dplyr to filter data.


Summary
-------

gghighlight package is a tool to highlight charactaristic data series among too many ones. Please try!

Bug reports or feature requests are welcome! -> https://github.com/yutannihilation/gghighlight/issues