---
title: "gghighlight: Highlight ggplot's Lines and Points with Predicates"
date: "2017-10-06"
categories: ["R"]
tags: ["gghighlight", "ggplot", "package"]
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

Suppose we have a data that has too many series like this:

```{r set-data}
set.seed(2)
d <- purrr::map_dfr(
  letters,
  ~ data.frame(idx = 1:400,
               value = cumsum(runif(400, -1, 1)),
               type = .,
               stringsAsFactors = FALSE))
```

For such data, it is almost impossible to identify a series by its colour as their differences are so subtle.

```{r plot}
library(ggplot2)

ggplot(d) +
  geom_line(aes(idx, value, colour = type))
```

So, I am motivated to filter data and map colour only on that, using dplyr:

```{r dplyr}
library(dplyr, warn.conflicts = FALSE)

d_filtered <- d %>%
  group_by(type) %>% 
  filter(max(value) > 20) %>%
  ungroup()

ggplot() +
  # draw the original data series with grey
  geom_line(aes(idx, value, group = type), data = d, colour = alpha("grey", 0.7)) +
  # colourise only the filtered data
  geom_line(aes(idx, value, colour = type), data = d_filtered)
```

But, what if I want to change the threshold in predicate (`max(.data$value) > 20`) and highlight other series as well? It's a bit tiresome to type all the code above again every time I replace `20` with some other value.

**gghighlight** package provides two functions to do this job. `gghighlight_line()` is the one for lines. The code equivalent to above (and more) can be this few lines:

```{r gghighlight-line-basic}
library(gghighlight)

gghighlight_line(d, aes(idx, value, colour = type), max(value) > 20)
```

As `gghighlight_*()` returns a ggplot object, it is fully customizable just as we usually do with ggplot2 like custom themes and facetting.

```{r gghighlight-theme}
library(ggplot2)

gghighlight_line(d, aes(idx, value, colour = type), max(value) > 20) +
  theme_minimal()
```

```{r gghighlight-facet}
gghighlight_line(d, aes(idx, value, colour = type), max(value) > 20) +
  facet_wrap(~ type)
```
