---
title: "dplyr Doesn't Provide Full Support For S4 (For Now?)"
date: "2018-02-12"
categories: ["R"]
tags: ["dplyr"]
---

I've seen sooo many (duplicated) issues on this topic were opened on dplyr's repo and lubridate's repo.

For example, you cannot `filter()` `interval`s. Consider the data below:

```{r interval-define}
library(lubridate)
library(dplyr, warn.conflicts = FALSE)

# some method of tibble() won't work for interval
d <- data.frame(
  i     = interval(ymd(10000101) + years(1:3 * 1000), ymd(10000102) + years(1:3 * 1000)),
  value = 1:3
)

d
```

Let's select the second row by `filter()`.

```{r interval-filter}
d %>% 
  filter(value == 2L)
```

Hmm, can you see something is wrong with the result? In case you don't notice yet, comparing with the result by base `subset()` may help:

```{r interval-subset}
subset(d, value == 2L)
```

As you see, the result should be `3000-01-01 UTC--3000-01-02 UTC`, whereas `filter()` returns `2000-01-01 UTC--2000-01-02 UTC`. Why? This is related to the structure of `interval` class. Let's examine them by `str()`:

```{r interval-str1}
str(d$i)
```

`interval` objects consist of 3 slots, 2 of which have the same length.
This means, to subset this vector, we need to subset `.Data` and `start` in the same way. But...


```{r interval-str2}
d %>% 
  filter(value == 2L) %>%
  str()
```

You can notice that it failed to subset `start`. In contrast, `subset()` properly subsets the slot as well:

```{r interval-str3}
str(subset(d, value == 2L))
```

This is because `subset()` dispatches the proper S4 method for `interval`, while dplyr fails to handle S4.

Of course, the maintainers are aware of this issue and seem to plan to address in long-awaited package, [vctrs](https://github.com/hadley/vctrs/).

* [Better support combining for non-base types 路 Issue #2432 路 tidyverse/dplyr](https://github.com/tidyverse/dplyr/issues/2432)
* [Support nonstandard representations? 路 Issue #27 路 hadley/vctrs](https://github.com/hadley/vctrs/issues/27)

So, apparently, the content of this post won't stay useful over time. But, for now, I feel this temporal "known issue" should be well-known, at least among those who suffers from this issue.

I hope this post will be outdated soon!