---
title: "Re-introduction to gghighlight: Highlight ggplot2 with Predicates"
date: "2018-06-16"
categories: ["R"]
tags: ["ggplot2", "gghighlight"]
editor_options: 
  chunk_output_type: console
---

Half a year ago, I've [introduced gghighlight](/post/2017-10-06-gghighlight/) package.
I didn't expect so much R peeps get interested in my package. Thanks for your attension!

But, please forget about this gghighlight; gghighlight has become far more powerful and simple!
So, let me re-introduce about gghighlight.

(Note that this version of gghighlight is not yet on CRAN at the time of this writing.)

## Motivation

### dplyr has `filter()`

What do you do when you explore a data that is too large to print?

```{r dplyr1, eval=FALSE}
library(dplyr, warn.conflicts = FALSE)

big_data %>%
  group_by(some_key) %>%
  summarise(some_agg = some_func(some_column))
# Opps, the result is too large!
```

dplyr's `filter()` is the Swiss army knife for this, which enables us to narrow down the data.
One nice thing of this function is that it can be inserted to any steps in the chain of `%>%`, so we don't need to rewrite the entire code.

```{r dplyr2, eval=FALSE}
big_data %>%
  # OK, let's filter the data
  filter(some_column > some_value) %>%
  group_by(some_key) %>%
  summarise(some_agg = some_func(some_column))
```

### ggplot2?

OK, good. But, what about ggplot2?

For a data that has too many series, it is almost impossible to identify a series by its colour as their differences are so subtle.

```{r ggplot2-simple, message=FALSE}
library(tidyverse)

set.seed(2)
d <- map_dfr(
  letters,
  ~ data.frame(
      idx = 1:400,
      value = cumsum(runif(400, -1, 1)),
      type = .,
      flag = sample(c(TRUE, FALSE), size = 400, replace = TRUE),
      stringsAsFactors = FALSE
    )
)

ggplot(d) +
  geom_line(aes(idx, value, colour = type))
```

Of course, I can use dplyr's `filter()` here as well.

```{r ggplot2-filter}
library(dplyr, warn.conflicts = FALSE)

d_filtered <- d %>%
  group_by(type) %>% 
  filter(max(value) > 20) %>%
  ungroup()

ggplot(d_filtered) +
  geom_line(aes(idx, value, colour = type))
```

But, it seems not so handy. For example, what if I want to change the threshold in predicate (`max(value) > 20`) and highlight other series as well? It’s a bit tiresome to type all the code above again every time I replace `20` with some other value...

So, I want `filter()` for ggplot2, which. This is my initial impulse to create gghighlight.

### Highlighting is better than filtering

In my understanding, one of the main purposes of visualization is to get the overview of a data.
In this sense, it may not be good to simply filter out the unmatched data because the plot loose its context then.
It's better to keep the unimportant data as grayed-out lines. Here comes the need for highlighting, like this:

```{r ggplot2-highlight-by-hand}
ggplot(d_filtered) +
  geom_line(aes(idx, value, group = type), data = d, colour = alpha("grey", 0.7)) +
  geom_line(aes(idx, value, colour = type))
```

This looks nicer!
So, now, my motivation has changed a bit; I want a function that highlights the important parts of a data, instead of filtering out the unimportant parts.

(If you are interested in the more details behind the idea of highlighting, you may find this post useful: [Anatomy of gghighlight](https://yutani.rbind.io/post/2018-06-03-anatomy-of-gghighlight/).)

## `gghighlight()`

Here is my answer, `gghighlight()`:

```{r gghighlight-simple}
library(gghighlight)

ggplot(d) +
  geom_line(aes(idx, value, colour = type)) +
  gghighlight(max(value) > 20)
```

Like filtering data with `filter()`, you can highlight the data by just adding `gghighlight()`.

Just like `filter()`, you can specify as many predicates as you like.
For example, the following code highlights the data that satisfies both `max(value) > 15` and `mean(flag) > 0.55`.

```{r gghighlight-two-conds}
ggplot(d) +
  geom_line(aes(idx, value, colour = type)) +
  gghighlight(max(value) > 15, mean(flag) > 0.55)
```

## Customization

As adding `gghighlight()` results in a ggplot object, it is fully customizable just as we usually do with ggplot2 like custom themes.

```{r gghighlight-theme}
ggplot(d) +
  geom_line(aes(idx, value, colour = type)) +
  gghighlight(max(value) > 19) +
  theme_minimal()
```

The plot also can be facetted:

```{r gghiglight-facet}
ggplot(d) +
  geom_line(aes(idx, value, colour = type)) +
  gghighlight(max(value) > 19) +
  theme_minimal() +
  facet_wrap(~ type)
```

## Geoms

`gghighlight()` can highlight almost every geoms. Here are some examples.

### Bar

`gghighlight()` can highlight bars.

```{r bar}
p <- ggplot(iris, aes(Sepal.Length, fill = Species)) +
  geom_histogram() +
  gghighlight()

p
```

You may wonder if this is really highlighted. Yes, it is.
But, the unhighlighted bars are all overwritten by the highlighted bars.
This seems not so useful, until you see the fecetted version:

```{r bar-wrap}
p + facet_wrap(~ Species)
```

### Point

As I explained in [Anatomy of gghighlight](https://yutani.rbind.io/post/2018-06-03-anatomy-of-gghighlight/), lines and points typically have different semantics (group-wise or not). But, in most cases, you don't need to be careful about the difference with `gghighlight()` because it automatically picks the right way of calculation.

```{r point}
set.seed(10)
d2 <- dplyr::sample_n(d, 20)

ggplot(d2, aes(idx, value)) +
  geom_point() +
  gghighlight(value > 0, label_key = type)
```

More precisely, `gghighlight()` takes the following strategy:

1. Calculate the group IDs from mapping.
    a. If `group` exists, use it.
    b. Otherwise, assign the group IDs based on the combination of the values of discrete variables.
2. If the group IDs exists, evaluate the predicates in a grouped manner.
3. If the group IDs doesn't exist or the grouped calculation fails, evaluate the predicates in an ungrouped manner.

Note that, in this case, `label_key = type` is needed to show labels because `gghighlights()` chooses a discrete variable from the mapping, but `aes(idx, value)` consists of only continuous variables.

### Sf

For the proof of gghiglight's capability, here's highlighted `geom_sf()`:

```{r sf}
nc <- sf::st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)

ggplot(nc) +
  geom_sf(aes(fill = AREA)) +
  gghighlight(grepl("^[A-C]", NAME)) +
  ggtitle("Polygons whose names start with A-C are highlighted!")
```

### (Exceptions)

I've written "`gghighlight()` can highlight **almost** every geoms." I mean, there are some exceptions that
gghighlight can not handle. But, I think I'm aware of only few of these. So, please let me know if you see
counter-intuitive results or errors via [GitHub](https://github.com/yutannihilation/gghighlight/issues) or [Twitter](https://twitter.com/yutannihilat_en) or SO!


## Non-logical predicate

To construct a predicate expression like bellow, we need to determine a threshold (in this example, `20`). But it is difficult to choose a nice one before we draw plots.

```{r predicate-example, eval=FALSE}
max(value) > 20
```

So, `gghighlight()` allows predicates that return numeric (or character) results. The values are used for sorting data and the top `max_highlight` of rows/groups are highlighted:

```{r numeric-highlight}
ggplot(d, aes(idx, value, colour = type)) +
  geom_line() +
  gghighlight(max(value), max_highlight = 5L)
```

## Backward-compatibility

`gghighlight_point()` and `gghighlight_line()` are here to stay for some time, but they will be deprecated in favor of `gghighlight()`.
The design of them was due to [the limitation of extendability of ggplot's `+` operator](https://yutani.rbind.io/post/2017-11-07-ggplot-add/), so it was not what it should be. Please consider using `gghighlight()` instead.

## Caveats

gghighlight is good to explore data by changing a threshlold little by little.
But, the internals are not so efficient, as it does almost the same calculation everytime you execute `gghighlight()`, which may get slower when it works with larger data. Consider doing this by using vanilla dplyr to filter data.

## Alternative

FWIW, here's a different approach of highlighting ggplot2 by [atusy](https://github.com/atusy/).
While my package modifies the clones of the existing layers, ggAtusy package modifies ggproto and create a new function that creates layers.
This approach seems clean and simple, whereas my code is full of tweaks and some kind of black magics...

* https://github.com/atusy/ggAtusy/blob/master/R/gghl.R

## Summary

gghighlight package has become cool. Please try!

Bug reports or feature requests are welcome! -> https://github.com/yutannihilation/gghighlight/issues