---
title: Quote While the Promise Is Hot!
author: ''
date: '2018-10-18'
slug: quote-while-the-promise-is-hot
categories:
  - R
tags:
  - tidyeval
  - tidyverse
description: ''
---

Suppose we want to quote `x` when `x` is not `NULL`. The naive implementation would be like below. Here, `y` is for comparison. Do you understand why `x` and `y` are quoted differently?

```{r quote}
quote_x_and_y <- function(x, y) {
  if (is.null(x)) {
    stop("x is NULL!", call. = FALSE)
  }
  
  x <- rlang::enquo(x)
  y <- rlang::enquo(y)
  
  list(x, y)
}

x <- y <- 1

quote_x_and_y(x, y)
```

This is because `x` is evaluated when `is.null()` is called before quoting, whereas `y` is intact. Lionel Henry, the tidyeval super hero, [answered my qustion on RStudio Community](https://community.rstudio.com/t/a-difference-between-tidyeval-and-substitute-when-the-promise-is-already-evaluated/15480/4?u=yutannihilation):

> A forced promise can no longer be captured correctly because it no longer carries an environment.

This means we must not touch arguments before quoting. Instead, quote first and check the expression inside quosure by `rlang::quo_is_*()`.


```{r quote2}
quote_x_and_y2 <- function(x, y) {
  x <- rlang::enquo(x)
  y <- rlang::enquo(y)
  
  if (rlang::quo_is_null(x)) {
    stop("x is NULL!", call. = FALSE)
  }
  
  list(x, y)
}

quote_x_and_y2(x, y)
```

For more complex checking, we may need to extract the expression from the quosure by `rlang::quo_get_expr()`.

```{r quote3}
quote_x_and_y_wont_stop <- function(x, y) {
  x <- rlang::enquo(x)
  y <- rlang::enquo(y)

  x_expr <- rlang::quo_get_expr(x)  
  if (rlang::call_name(x) %in% "stop") {
    message("Nothing can stop me!\n")
  }
  
  list(x, y)
}

quote_x_and_y_wont_stop(stop("foo"), "bar")
```

Anyway, keep in mind to use `enquo()` (or `ensym()`) at the very beginning of the function. Quote while the promise is hot.