---
title: "Unofficial Introduction To extendr (1): Your First R Package With Rust"
description: |
  Integrate R and Rust with extendr.
author:
  - name: Hiroaki Yutani
    url: {}
categories:
  - Rust
  - extendr
date: 06-06-2021
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
    self_contained: false
creative_commons: CC BY
---


[extendr](https://github.com/extendr/extendr) is a project that provides an
interface between R and Rust. While I'm also a member of this project,
my contributions were small and I'm still a newbie to Rust, so let me add
"Unofficial" to the title at the moment :P

For the full introduction, including the motivation to use Rust, please watch
this great presentation by Claus Wilke:

<iframe width="560" height="315" src="https://www.youtube.com/embed/EX7YG2pmcC8" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

The talk covers almost everything (history of the project, getting started with Rust,
and how to compile a Rust function into an R function), so, instead of providing
an overview, this post will focus on the one missing topic, how to write an R
package using (r)extendr.

## Rust crates and R packages under extendr

That said, to avoid confusion, probably I need to write a minimal introduction
about the crates and packages we'll look into...
Let's take a look briefly.

### libR-sys (Rust crate)

[libR-sys](https://github.com/extendr/libR-sys) is low-level bindings to R's C
API, and powers extendr. This is not what the ordinal users use directly,
so we can forget this name for now.

### extendr (Rust crate)

[extendr](https://github.com/extendr/extendr) is a user-friendly Rust interface
to R. To be precise, "extendr" itself is not a crate's name; it's `extendr-api`
that we'll actually use. There are some more crates that might be needed in some
use cases, but let's forget them for now.

### rextendr (R package)

[**r**extendr](https://github.com/extendr/rextendr) is an R package. This provides
two types of functions:

1. Functions for compiling and running Rust code on the fly (e.g. [`rust_function()`](https://extendr.github.io/rextendr/reference/rust_source.html))
2. Functions for developing R package using extendr

What we'll use in this post is type 2.

## Setup

### Rust

First of all, you need Rust.

If you are using macOS or Linux, you can follow the [official guide](https://www.rust-lang.org/tools/install). That's all.

If you are using Windows, extendr requires you to install the latest Rtools, and
use the following toolchains:

``` sh
rustup default stable-msvc
rustup target add x86_64-pc-windows-gnu  # 64-bit
rustup target add i686-pc-windows-gnu    # 32-bit
```

For the details, please refer to the installation instructions on [libR-sys's repo](https://github.com/extendr/libR-sys/#installation).

### rextendr

rextendr is not on CRAN at the time of writing this blog post, so please install it from GitHub.

```{r install-rextendr, eval=FALSE}
remotes::install_github("extendr/rextendr")
```

## Create a template package

Creating an R package with extendr is very easy with usethis and rextendr.

First, create an R package by using `usethis::create_package()` as usual.

```{r create-package, eval=FALSE}
usethis::create_package("path/to/my1stextendrpkg")
```

Then, create the scaffolding to use extendr. This can be done with `rextendr::use_extendr()`.

```{r use-rextendr, eval=FALSE}
rextendr::use_extendr()
```

```
✓ Creating src/rust/src.
✓ Setting active project to 'path/to/my1stextendrpkg'
✓ Writing 'src/entrypoint.c'
✓ Writing 'src/Makevars'
✓ Writing 'src/Makevars.win'
✓ Writing 'src/.gitignore'
✓ Writing src/rust/Cargo.toml.
✓ Writing 'src/rust/src/lib.rs'
✓ Writing 'R/extendr-wrappers.R'
✓ Finished configuring extendr for package my1stextendrpkg.
• Please update the system requirement in DESCRIPTION file.
• Please run `rextendr::document()` for changes to take effect.
```

Done! Now we are just one step away (as the message says, we need to run `rextendr::document()`) from calling Rust fucntions from R.
But, before moving forward, let's look at the files added.

## Package structure

The below files are the ones `rextendr::use_extendr()` added.

```
.
├── R
│   └── extendr-wrappers.R
...
└── src
    ├── Makevars
    ├── Makevars.win
    ├── entrypoint.c
    └── rust
        ├── Cargo.toml
        └── src
            └── lib.rs
```

* **`R/extendr-wrappers.R`**: This file contains auto-generated R functions from Rust code. We don't modify this by hand.
* **`src/Makevars`**, **`src/Makevars.win`**: This hooks `cargo build` at the installation of the R package. In most of the cases, we don't edit these.
* **`src/entrypoint.c`**: This is needed to avoid the linker removing the static library. In 99.9% of the cases, we don't edit this (except for changing the crate name).
* **`src/rust/`**: Rust code of a crate using extendr-api. This is where we mainly write code.

So, in short, what we should really look at is only these two files:

### `src/rust/Cargo.toml`


``` toml
[package]
name = 'my1stextendrpkg'
version = '0.1.0'
edition = '2018'

[lib]
crate-type = [ 'staticlib' ]

[dependencies]
extendr-api = '*'
```

The create name is the same name as the R package's name by default. You can change this,
but it might be a bit tired to tweak other files accordingly, so I recommend leaving this.

To try the dev version of the extendr, you can modify the last line to

```toml
extendr-api = { git = 'https://github.com/extendr/extendr' }
```

### `src/rust/src/lib.rs`

``` rs
use extendr_api::prelude::*;

/// Return string `"Hello world!"` to R.
/// @export
#[extendr]
fn hello_world() -> &'static str {
    "Hello world!"
}

// Macro to generate exports.
// This ensures exported functions are registered with R.
// See corresponding C code in `entrypoint.c`.
extendr_module! {
    mod my1stextendrpkg;
    fn hello_world;
}
```

Let's explain this part by part.

The first line `use extendr_api::prelude::*;` loads functions used frequently.

Next, your eyes might notice the `/` are repeated 3 times, while the usual
Rust comment requires only twice (i.e. `//`). These are treated as roxygen comments
and copied to the auto-generated R code. This is analogous to Rcpp/cpp11's `//'`.

``` rs
/// Return string `"Hello world!"` to R.
/// @export
```

The next line is the core of `extendr`'s mechanism. If the function is marked with this
macro, the corresponding R function will be generated automatically (I'll explain
the detail later). This is analogous to Rcpp's `[[Rcpp::export]]` and cpp11's `[[cpp11::register]]`.

``` rs
#[extendr]
```

The last 3 lines are the macro for generating exports, as the comment explains.
If we implement another function than `hello_world`, it needs to be listed here
as well as marking it with `#[extendr]` macro.

``` rs
extendr_module! {
    mod my1stextendrpkg;
    fn hello_world;
}
```

## Compile and use the package

### Compile

Compiling Rust code into R functions is as easy as this one command:

```{r build-and-use, eval=FALSE}
rextendr::document()
```

```
✓ Saving changes in the open files.
ℹ Generating extendr wrapper functions for package: my1stextendrpkg.
! No library found at src/my1stextendrpkg.so, recompilation is required.
Re-compiling my1stextendrpkg
─  installing *source* package ‘my1stextendrpkg’ ... (347ms)
   ** using staged installation
   ** libs
   rm -Rf my1stextendrpkg.so ./rust/target/release/libmy1stextendrpkg.a entrypoint.o
   gcc -std=gnu99 -I"/usr/share/R/include" -DNDEBUG      -fpic  -g -O2 -fdebug-prefix-map=/build/r-base-tbZjLv/r-base-4.1.0=. -fstack-protector-strong -Wformat -Werror=format-security -Wdate-time -D_FORTIFY_SOURCE=2 -g  -UNDEBUG -Wall -pedantic -g -O0 -fdiagnostics-color=always -c entrypoint.c -o entrypoint.o
   cargo build --lib --release --manifest-path=./rust/Cargo.toml
       Updating crates.io index
      Compiling proc-macro2 v1.0.27
      Compiling unicode-xid v0.2.2
      Compiling libR-sys v0.2.1
      Compiling syn v1.0.72
      Compiling extendr-engine v0.2.0
      Compiling lazy_static v1.4.0
      Compiling quote v1.0.9
      Compiling extendr-macros v0.2.0
      Compiling extendr-api v0.2.0
      Compiling my1stextendrpkg v0.1.0 (path/to/my1stextendrpkg/src/rust)
       Finished release [optimized] target(s) in 19.05s
   gcc -std=gnu99 -shared -L/usr/lib/R/lib -Wl,-Bsymbolic-functions -Wl,-z,relro -o my1stextendrpkg.so entrypoint.o -L./rust/target/release -lmy1stextendrpkg -L/usr/lib/R/lib -lR
   installing to /tmp/RtmpfMcL08/devtools_install_e2d6351b843c/00LOCK-my1stextendrpkg/00new/my1stextendrpkg/libs
   ** checking absolute paths in shared objects and dynamic libraries
─  DONE (my1stextendrpkg)
✓ Writing 'R/extendr-wrappers.R'.
ℹ Updating my1stextendrpkg documentation
ℹ Loading my1stextendrpkg
Writing NAMESPACE
Writing NAMESPACE
Writing hello_world.Rd
```

You might wonder why compilation is triggered while the function name is just `document()`.
Well, this is because the compilation is actually needed to generate document
from Rust code. This is consistent with `devtools::document()`'s behavior for C/C++ codes[^hook].

Anyway, by doing above, the following files are updated or generated:

```
.
...
├── NAMESPACE                       ----------(4)
├── R
│   └── extendr-wrappers.R          ----------(3)
├── man
│   └── hello_world.Rd              ----------(4)
└── src
    ├── my1stextendrpkg.so          ----------(2)
    └── rust
        └── target
            └── release
                ├── libmy1stextendrpkg.a   ---(1)
                ...
```

1. **`src/rust/target/release/libmy1stextendrpkg.a`** (the extension depends on the OS): 
This is the static library built from Rust code. This will be then used for compiling
shared library `my1stextendrpkg.so`.
2. **`src/my1stextendrpkg.so`** (the extension depends on the OS): This is the
shared object that is actually called from R.
3. **`R/extendr-wrappers.R`**: The auto-generated R functions, including roxygen
comments, goes to this file. The roxygen comments are accordingly converted into
Rd files and `NAMESPACE`.
4. **`man/`**, **`NAMESPACE`**: These are generated from roxygen comments.

### Load and use

As all things are done by `rexetndr::document()` already, we can just load it 
(or install it if you want) and call the function.

```{r load1, eval=FALSE}
devtools::load_all(".")

hello_world()
```

```{r load2, echo=FALSE}
devtools::load_all("~/repo/my1stextendrpkg/")

hello_world()
```

Achievement unlocked, you called a Rust function from R!

## Rust code vs generated R code

We don't open `R/extendr-wrappers.R` yet. While we never edit this file by hand,
it might be good to know what R code is generated from a Rust code. Here it is:


``` r
# Generated by extendr: Do not edit by hand
#
# This file was created with the following call:
#   .Call("wrap__make_my1stextendrpkg_wrappers", use_symbols = TRUE, package_name = "my1stextendrpkg")

#' @docType package
#' @usage NULL
#' @useDynLib my1stextendrpkg, .registration = TRUE
NULL

#' Return string `"Hello world!"` to R.
#' @export
hello_world <- function() .Call(wrap__hello_world)
```

`.Call("wrap__make_my1stextendrpkg_wrappers", use_symbols = ...` is
what was actually done inside `rextendr::document()`.

A section of `@docType package` is needed to generate `useDynLib(my1stextendrpkg, .registration = TRUE)` entry
in `NAMESPACE`.

The last section is for `hello_world()`. We can see the roxygen comments are copied
to here. As the Rust function `hello_world()` has no arguments so this R function
also has no arguments. If the function is like this,

``` rs
fn add(x: i32, y: i32) -> i32 {
    x + y
}
```

then the generated function also has arguments like this:

``` r
add <- function(x, y) .Call(wrap__add, x, y)
```

## Implement a new Rust function

Now that we roughly figured out how extendr works (hopefully!), let's implment
a new Rust function. The development flow would be:

1. Modify `src/rust/src/lib.rs`
2. Run `rextendr::document()`
3. Run `devtools::load_all(".")` and test the function

As an exercise, let's add `add(i32, i32)` I showed above.

### 1. Modify `src/rust/src/lib.rs`

Add the function with `@export`.

``` rs
/// @export
#[extendr]
fn add(x: i32, y: i32) -> i32 {
    x + y
}
```

Don't forget to add the function to `extendr_module!`.

``` rs
extendr_module! {
    mod my1stextendrpkg;
    fn hello_world;
    fn add;
}
```

### 2. Run `rextendr::document()`

Just run the command:

```{r doc2, eval=FALSE}
rextendr::document()
```

### 3. Run `devtools::load_all(".")` and test the function

Ditto.

```{r load3, eval=FALSE}
devtools::load_all(".")

add(1L, 2L)
```

```{r load4, echo=FALSE}
devtools::load_all("~/repo/my1stextendrpkg/")

add(1L, 2L)
```

Achievement unlocked, you called a Rust function you implemented from R!

## What's not covered in this post

I hope this post illustrates how easy it is to get started with extendr to create an R package.
But, if you play with the function, you might get wondered about the topics that this post doesn't cover.

For example, while the signature is `i32` (i.e. integer), the function also accepts numeric.
What's the rule behind this coercion? (Confession: I need to study to answer this question...)

```{r add_real}
add(1, 2)
```

Another question might be how to handle vectors. This function accepts only length-one
vectors. Otherwise, it errors. (Spoiler: This is very simple; we can use `Vec<_>`. But, let me leave this topic to the next post...)

```{r add_real2, error=TRUE}
add(1:2, 2:3)
```

I'll try explaining these in the next post. Stay tuned...

Until then, you can ask questions on extendr's Discord, which you can find on <https://github.com/extendr/extendr#contributing> :)


[^hook]: The current mechanism under `devtools::document()` doesn't have extensible mechanism
for other languages than C/C++ (c.f. [r-lib/pkgbuild#115](https://github.com/r-lib/pkgbuild/issues/115)), so we needed to have our own one.
